You can combine templates with `&&` / `||`:

```{r}
vet(numeric(1L) || NULL, NULL)
vet(numeric(1L) || NULL, 42)
vet(numeric(1L) || NULL, "foo")
```

Templates only check structure.  When you need to check values use `.` to
refer to the object:

```{r}
vet(numeric(1L) && . > 0, -42)  # strictly positive scalar numeric
vet(numeric(1L) && . > 0, 42)
```

You can compose vetting expressions as language objects and combine them:

```{r}
scalar.num.pos <- quote(numeric(1L) && . > 0)
foo.or.bar <- quote(character(1L) && . %in% c('foo', 'bar'))
vet.exp <- quote(scalar.num.pos || foo.or.bar)

vet(vet.exp, 42)
vet(vet.exp, "foo")
vet(vet.exp, "baz")
```

There are a number of predefined vetting tokens you can use in your
vetting expressions:

```{r}
vet(NUM.POS, -runif(5))    # positive numeric
vet(LGL.1, NA)             # TRUE or FALSE
```

See `?vet_token` for a full listing, and for instructions on how to define your
own tokens with custom error messages.

Vetting expressions are designed to be intuitive to use, but their
implementation is complex.  We recommend you look at `example(vet)` for usage
ideas, or at the ["Non Standard Evaluation" section of the vignette][3] for the
gory details.

