<!-- README.md is generated from README.Rmd. Please edit that file 

library(rmarkdown)
render('README.Rmd', output_format=html_vignette(css='vignettes/styles.css'))
rmarkdown::render('README.Rmd', output_format=rmarkdown::md_document())

-->
```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  fig.path = "README-",
  error = TRUE
)
library(vetr)
```

# vetr

[![R build status](https://github.com/brodieG/vetr/workflows/R-CMD-check/badge.svg)](https://github.com/brodieG/vetr/actions)
[![](https://codecov.io/github/brodieG/vetr/coverage.svg?branch=master)](https://app.codecov.io/gh/brodieG/vetr?branch=master)
[![](https://codecov.io/github/brodieG/vetr/coverage.svg?branch=master)](https://codecov.io/github/brodieG/vetr?branch=master)
[![](http://www.r-pkg.org/badges/version/vetr)](https://cran.r-project.org/package=vetr)
[![Project Status: WIP - Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![Dependencies direct/recursive](https://tinyverse.netlify.app/badge/vetr)]( https://tinyverse.netlify.app/)

```{r child='vignettes/rmdhunks/trust-but-verify.Rmd'}
```

```{r child='vignettes/rmdhunks/declarative-checks.Rmd'}
```

## Vetting Expressions

```{r child='vignettes/rmdhunks/vetting-expressions.Rmd'}
```

## `vetr` in Functions

If you are vetting function inputs, you can use the `vetr` function, which works
just like `vet` except that it is streamlined for use within functions:

```{r}
fun <- function(x, y) {
  vetr(numeric(1L), logical(1L))
  TRUE   # do work...
}
fun(1:2, "foo")
fun(1, "foo")
```

`vetr` automatically matches the vetting expressions to the corresponding
arguments and fetches the argument values from the function environment.

See [vignette][4] for additional details on how the `vetr` function works.

## Additional Documentation

* [`vetr` vignette][1], `?vet`, `?vetr`, `example(vet)`, `example(vetr)`.
* [`alike` vignette][2], `?alike`, and `example(alike)` for discussion of
  templates.
* A survey of [parameter validation functions][5].

## Development Status

`vetr` is still in development, although most of the features are considered
mature.  The most likely area of change is the treatment of function and
language templates (e.g.  `alike(sum, max)`), and more flexible treatment of
list templates (e.g. in future lists may be allowed to be different lengths so
long as every named element in the template exists in the object).

## Installation

This package is available on CRAN:

```{r eval=FALSE}
install.packages('vetr')
```

It has no runtime dependencies.

For the development version use
`remotes::install_github('brodieg/vetr@development')` or:

```{r eval=FALSE}
f.dl <- tempfile()
f.uz <- tempfile()
github.url <- 'https://github.com/brodieG/vetr/archive/development.zip'
download.file(github.url, f.dl)
unzip(f.dl, exdir=f.uz)
install.packages(file.path(f.uz, 'vetr-development'), repos=NULL, type='source')
unlink(c(f.dl, f.uz))
```

The master branch typically mirrors CRAN and should be stable.

## Alternatives

There are many alternatives available to `vetr`.  We do a survey of the
following in our [parameter validation functions][5] review:

```{r child='vignettes/rmdhunks/related-packages.Rmd'}
```

## Acknowledgments

Thank you to:

* R Core for developing and maintaining such a wonderful language.
* CRAN maintainers, for patiently shepherding packages onto CRAN and maintaining
  the repository, and Uwe Ligges in particular for maintaining
  [Winbuilder](https://win-builder.r-project.org/).
* Users and others who have reported bugs and/or helped contribute fixes (see
  NEWS.md).
* Tomas Kalibera for [rchk](https://github.com/kalibera/rchk) and rcnst to help
  detect errors in compiled code, and in particular for his infinite patience in
  helping me resolve the issues he identified for me.
* [Jim Hester](https://github.com/jimhester) because
  [covr](https://cran.r-project.org/package=covr) rocks.
* [Dirk Eddelbuettel](https://github.com/eddelbuettel) and [Carl
  Boettiger](https://github.com/cboettig) for the
  [rocker](https://github.com/rocker-org/rocker) project, and [Gábor
  Csárdi](https://github.com/gaborcsardi) and the
  [R-consortium](https://www.r-consortium.org/) for
  [Rhub](https://github.com/r-hub), without which testing bugs on R-devel and
  other platforms would be a nightmare.
* [Winston Chang](https://github.com/wch) for the
  [r-debug](https://hub.docker.com/r/wch1/r-debug/) docker container, in
  particular because of the valgrind level 2 instrumented version of R.
* [Hadley Wickham](https://github.com/hadley/) and [Peter
  Danenberg](https://github.com/klutometis) for
  [roxygen2](https://cran.r-project.org/package=roxygen2).
* [Yihui Xie](https://github.com/yihui) for
  [knitr](https://cran.r-project.org/package=knitr) and  [J.J.
  Allaire](https://github.com/jjallaire) etal for
  [rmarkdown](https://cran.r-project.org/package=rmarkdown), and by extension
  John MacFarlane for [pandoc](https://pandoc.org/).
* [Michel Lang](https://github.com/mllg) for pushing me to implement `all_bw` to
  compete with his own package [`checkmate`](https://cran.r-project.org/package=checkmate).
* [Eugene Ha](https://github.com/egnha) for pointing me to several other
  relevant packages, which in turn led to the [survey of related packages][5].
* Stefan M. Bache for the idea of having a function for testing objects directly
  (originally `vetr` only worked with function arguments), which I took from
  ensurer.
* Olaf Mersmann for
  [microbenchmark](https://cran.r-project.org/package=microbenchmark), because
  microsecond matter, and [Joshua Ulrich](https://github.com/joshuaulrich) for
  making it lightweight.
* All open source developers out there that make their work freely available
  for others to use.
* [Github](https://github.com/), [Codecov](https://about.codecov.io/),
  [Vagrant](https://www.vagrantup.com/), [Docker](https://www.docker.com/),
  [Ubuntu](https://ubuntu.com/), [Brew](https://brew.sh/) for providing
  infrastructure that greatly simplifies open source development.
* [Free Software Foundation](https://www.fsf.org/) for developing the GPL
  license and promotion of the free software movement.

## About the Author

Brodie Gaslam is a hobbyist programmer based on the US East Coast.

[1]: https://cran.r-project.org/package=vetr/vignettes/vetr.html
[2]: https://cran.r-project.org/package=vetr/vignettes/alike.html
[3]: https://cran.r-project.org/package=vetr/vignettes/vetr.html#non-standard-evaluation
[4]: https://cran.r-project.org/package=vetr/vignettes/vetr.html#in-functions
[5]: http://htmlpreview.github.io/?https://github.com/brodieG/vetr/blob/master/extra/compare.html
