<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  fig.path = "README-"
)
```

# vetr - Trust, but Verify

[![](https://travis-ci.org/brodieG/vetr.svg?branch=master)](https://travis-ci.org/brodieG/vetr)
[![](https://codecov.io/github/brodieG/vetr/coverage.svg?branch=master)](https://codecov.io/github/brodieG/vetr?branch=master)
[![](http://www.r-pkg.org/badges/version/vetr)](https://cran.r-project.org/package=vetr)
[![Project Status: WIP - Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](http://www.repostatus.org/badges/latest/wip.svg)](http://www.repostatus.org/#wip)

```{r child='vignettes/rmdhunks/trust-but-verify.Rmd'}
```

```{r child='vignettes/rmdhunks/declarative-checks.Rmd'}
```

## Vetting Expressions

```{r child='vignettes/rmdhunks/vetting-expressions.Rmd'}
```

## `vetr` in Functions

If you are vetting function inputs, you can use the `vetr` function, which works
just like `vet` except that it is streamlined for use within functions:

```{r}
fun <- function(x, y) {
  vetr(numeric(1L), logical(1L))
  TRUE   # do work...
}
fun(1:2, "foo")
fun(1, "foo")
```

`vetr` automatically matches the vetting expressions to the corresponding
arguments and fetches the argument values from the function environment.

See [vignette][1] for additional details on how the `vetr` function works.

## Additional Documentation

* [`vetr` vignette][1], `?vetr`, `example(vetr)`
* [`alike` vignette][2], `?alike`, and `example(alike)` for discussion of
  templates

## Development Status

`vetr` is still in development, although most of the features are considered
mature.  The most likely area of change is the treatment of function and
language templates (e.g.  `alike(sum, max)`), and more flexible treatment of
list templates (e.g. in future lists may be allowed to be different lengths so
long as every named element in the template exists in the object).

## Installation

`vetr` will shortly be available on CRAN.  In the meantime, you can get it from
github:

```{r, eval=FALSE}
# install.packages('devtools')
devtools::install_github('brodieg/vetr@development')
```

## Related Packages

<ul>
  <li><a href='https://github.com/egnha/valaddin'>valaddin</a> by Eugene Ha (see
  <a
  href='http://htmlpreview.github.io/?https://github.com/brodieG/vetr/blob/development/inst/doc/vetr.html#valaddin'>vignette</a>
  for a more detailed comparison) has very similar objectives to `vetr`
```{r child='vignettes/rmdhunks/related-packages.Rmd'}
```
</ul>

## Acknowledgments

Thank you to:

* R Core for developing and maintaining such a wonderfully language.
* CRAN maintainers, for patiently shepherding packages onto CRAN and maintaining
  the repository, and Uwe Ligges in particular for maintaining
  [Winbuilder](http://win-builder.r-project.org/).
* [Jim Hester](https://github.com/jimhester) because
  [covr](https://cran.r-project.org/package=covr) rocks.
* [Dirk Eddelbuettel](https://github.com/eddelbuettel) and [Carl
  Boettiger](https://github.com/cboettig) for the
  [rocker](https://github.com/rocker-org/rocker) project, and [Gábor
  Csárdi](https://github.com/gaborcsardi) and the R-consortium for
  [Rhub](https://github.com/r-hub/rhub), without which testing bugs on R-devel
  and other platforms would be a nightmare.
* [Yihui Xie](https://github.com/yihui) for
  [knitr](https://cran.r-project.org/package=knitr) and  [J.J.
  Allaire](https://github.com/jjallaire) etal for
  [rmarkdown](https://cran.r-project.org/package=rmarkdown), and by extension
  John MacFarlane for [pandoc](http://pandoc.org/).
* Stefan M. Bache for the idea of having a function for testing objects directly
  (originally `vetr` only worked with function arguments), which I took from
  ensurer.
* Hadley Wickham for [devtools](https://cran.r-project.org/package=devtools),
  and for pointing me to Stefan M. Bache's ensurer package.
* Olaf Mersmann for [microbenchmark](https://cran.r-project.org/package=microbenchmark), because microsecond matter.
* All open source developers out there that make their work freely available
  for others to use.
* [Github](https://github.com/), [Travis-CI](https://travis-ci.org/),
  [Codecov](https://codecov.io/), [Vagrant](https://www.vagrantup.com/),
  [Docker](https://www.docker.com/), [Ubuntu](https://www.ubuntu.com/),
  [Brew](https://brew.sh/) for providing infrastructure that greatly simplifies
  open source development.
* [Free Software Foundation](http://fsf.org/) for developing the GPL license and
  promotion of the free software movement.

## About the Author

Brodie Gaslam is a hobbyist programmer based on the US East Coast.

[1]: http://htmlpreview.github.io/?https://github.com/brodieG/vetr/blob/development/inst/doc/vetr.html
[2]: http://htmlpreview.github.io/?https://github.com/brodieG/vetr/blob/development/inst/doc/alike.html
[3]: http://htmlpreview.github.io/?https://github.com/brodieG/vetr/blob/development/inst/doc/vetr.html#non-standard-evaluation
