<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  fig.path = "README-",
  error = TRUE,
  comment = NA
)
library(vetr)
```

# vetr

[![](https://travis-ci.org/brodieG/vetr.svg?branch=master)](https://travis-ci.org/brodieG/vetr)
[![](https://codecov.io/github/brodieG/vetr/coverage.svg?branch=master)](https://codecov.io/github/brodieG/vetr?branch=master)
[![](http://www.r-pkg.org/badges/version/vetr)](https://cran.r-project.org/package=vetr)
[![Project Status: WIP - Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](http://www.repostatus.org/badges/latest/wip.svg)](http://www.repostatus.org/#wip)

```{r child='vignettes/rmdhunks/trust-but-verify.Rmd'}
```

```{r child='vignettes/rmdhunks/declarative-checks.Rmd'}
```

## Vetting Expressions

```{r child='vignettes/rmdhunks/vetting-expressions.Rmd'}
```

## `vetr` in Functions

If you are vetting function inputs, you can use the `vetr` function, which works
just like `vet` except that it is streamlined for use within functions:

```{r}
fun <- function(x, y) {
  vetr(numeric(1L), logical(1L))
  TRUE   # do work...
}
fun(1:2, "foo")
fun(1, "foo")
```

`vetr` automatically matches the vetting expressions to the corresponding
arguments and fetches the argument values from the function environment.

See [vignette][4] for additional details on how the `vetr` function works.

## Additional Documentation

* [`vetr` vignette][1], `?vet`, `?vetr`, `example(vet)`, `example(vetr)`.
* [`alike` vignette][2], `?alike`, and `example(alike)` for discussion of
  templates.
* A survey of [parameter validation functions][5].

## Development Status

`vetr` is still in development, although most of the features are considered
mature.  The most likely area of change is the treatment of function and
language templates (e.g.  `alike(sum, max)`), and more flexible treatment of
list templates (e.g. in future lists may be allowed to be different lengths so
long as every named element in the template exists in the object).

## Installation

```{r, eval=FALSE}
install.packages('vetr')
```

Or for the development version:

```{r, eval=FALSE}
# install.packages('devtools')
devtools::install_github('brodieg/vetr@development')
```

## Alternatives

There are many alternatives available to `vetr`.  We do a survey of the
following in our [parameter validation functions][5] review:

```{r child='vignettes/rmdhunks/related-packages.Rmd'}
```

## Acknowledgments

Thank you to:

* R Core for developing and maintaining such a wonderful language.
* CRAN maintainers, for patiently shepherding packages onto CRAN and maintaining
  the repository, and Uwe Ligges in particular for maintaining
  [Winbuilder](http://win-builder.r-project.org/).
* [Jim Hester](https://github.com/jimhester) because
  [covr](https://cran.r-project.org/package=covr) rocks.
* [Dirk Eddelbuettel](https://github.com/eddelbuettel) and [Carl
  Boettiger](https://github.com/cboettig) for the
  [rocker](https://github.com/rocker-org/rocker) project, and [Gábor
  Csárdi](https://github.com/gaborcsardi) and the R-consortium for
  [Rhub](https://github.com/r-hub/rhub), without which testing bugs on R-devel
  and other platforms would be a nightmare.
* [Yihui Xie](https://github.com/yihui) for
  [knitr](https://cran.r-project.org/package=knitr) and  [J.J.
  Allaire](https://github.com/jjallaire) etal for
  [rmarkdown](https://cran.r-project.org/package=rmarkdown), and by extension
  John MacFarlane for [pandoc](http://pandoc.org/).
* [Michel Lang]() for pushing me to implement `all_bw` to compete with his own
  package [`checkmate`](https://github.com/mllg/checkmate).
* [Eugene Ha]() for pointing me to several other relevant packages, which in
  turn led to the [survey of related packages][5].
* Stefan M. Bache for the idea of having a function for testing objects directly
  (originally `vetr` only worked with function arguments), which I took from
  ensurer.
* Hadley Wickham for [devtools](https://cran.r-project.org/package=devtools),
  and for pointing me to Stefan M. Bache's ensurer package.
* [Markus Kuhn](http://www.cl.cam.ac.uk/~mgk25/) for [UTF-8 stress test
  data](http://www.cl.cam.ac.uk/~mgk25/ucs/examples/UTF-8-test.txt)
* Olaf Mersmann for
  [microbenchmark](https://cran.r-project.org/package=microbenchmark), because
  microsecond matter.
* All open source developers out there that make their work freely available
  for others to use.
* [Github](https://github.com/), [Travis-CI](https://travis-ci.org/),
  [Codecov](https://codecov.io/), [Vagrant](https://www.vagrantup.com/),
  [Docker](https://www.docker.com/), [Ubuntu](https://www.ubuntu.com/),
  [Brew](https://brew.sh/) for providing infrastructure that greatly simplifies
  open source development.
* [Free Software Foundation](http://fsf.org/) for developing the GPL license and
  promotion of the free software movement.

## About the Author

Brodie Gaslam is a hobbyist programmer based on the US East Coast.

[1]: http://htmlpreview.github.io/?https://github.com/brodieG/vetr/blob/master/inst/doc/vetr.html
[2]: http://htmlpreview.github.io/?https://github.com/brodieG/vetr/blob/master/inst/doc/alike.html
[3]: http://htmlpreview.github.io/?https://github.com/brodieG/vetr/blob/master/inst/doc/vetr.html#non-standard-evaluation
[4]: http://htmlpreview.github.io/?https://github.com/brodieG/vetr/blob/master/inst/doc/vetr.html#in-functions
[5]: http://htmlpreview.github.io/?https://github.com/brodieG/vetr/blob/development/extra/compare.html
