---
title: "Parameter Validation Functions"
author: "Brodie Gaslam"
output:
    rmarkdown::html_vignette:
        toc: true
        css: ../vignettes/styles.css

vignette: >
  %\VignetteIndexEntry{vetr}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(error=TRUE)
library(vetr)
library(assertive)
library(assertthat)
library(ensurer)
library(checkmate)
library(valaddin)
```

## Prevent Garbage In

Most functions expect their parameters to be of specific types, structures, or
values.  If we allow functions to operate on parameters that do not conform to
the expectations then we could get in the best case difficult to debug errors
from the bowels of the function, and in the worst silent undefined behavior.

Systematically vetting function parameters is as tedious as it is important, but
fortunately many packages exist to assist with the process.  We will review
those packages and assess them across these dimensions:

* Ease of use for the function developer incorporating parameter checks
* Computing overhead
* Ease of use for end users

We will review the following packages / functions:

* `stopifnot` by R Core, evaluates test expressions and `stops` if they return
  anything other than all TRUE vectors.
* [`vetr`](https://github.com/brodieG/vetr) by Yours Truly, augments test
  expressions with templates for object structure verification.  This allows for
  a succinct and extensible declarative syntax.  The package also emphasizes
  informative error messages and speed.
* [`asserthat`](https://github.com/hadley/assertthat) by Hadley Wickham, a
  drop-in replacement for `stopifnot` with more helpful error messages for
  checks based on the base `is.*` functions (e.g. `is.character`) and some
  additional helper functions for common checks.
* [`assertive`](https://www.r-pkg.org/pkg/assertive) by Richie Cotton, a
  large library of checking functions for most common parameter checks with
  expressive error messages.
* [`checkmate`](https://github.com/mllg/checkmate) by Michael Lang, a library of
  flexible checking functions for most common types of objects, with an emphasis
  on speed and conciseness, particularly for vectors.
* [`ensurer`](https://github.com/smbache/ensurer) by Stefan M. Bache, a
  framework for constructing re-usable and extensible check contracts, with a
  flexible reporting channel, a mechanism for applying check contracts to
  function _outputs_, and an experimental interface for embedding the checking
  contracts into functions.

Not all of these packages are specifically designed to test function parameters,
but they can all be used that way.  We will also focus on "out of the box"
functionality.  For example, most of these packages support custom user supplied
error messages, but we evaluate only the error messages produced without
additional user intervention.  We also ignore functionality that is not
specifically used for function parameter checking.

The following packages also perform related tasks, although we do not review
them here:

* [`valaddin`](https://github.com/egnha/valaddin) by Eugene Ha, a framework for
  augmenting existing functions with validation contracts.  Currently the
  package is undergoing a major overhaul so we will add it to the comparison
  once the new release (v0.3.0) is out.
* [`validate`](https://github.com/data-cleaning/validate) by Mark van
  der Loo and Edwin de Jonge, with a primary focus on validating data in data
  frames and similar data structures
* [`assertr`](https://github.com/ropensci/assertr) by Tony Fischetti, also
  focused on data validation in data frames and similar structures
* [`types`](https://github.com/jimhester/types) by Jim Hester, which implements
  but does not enforce type hinting
* [`argufy`](https://github.com/gaborcsardi/argufy) by Gábor Csárdi, which
  implements parameter validation via roxygen tags (not released to CRAN)

## Summary

Every package is compared to `stopifnot` across several dimensions.  The only
objective measurements here are the timings  Everything else is my opinion
which you can safely assume is biased in favor of my own package `vetr`.

Scores are in relation to `stopifnot`, with positive numbers meaning the package
performs better than `stopifnot`, and negative numbers that it performs worse.

In this section we will review the high level summary of our comparisons.  For
specific details see the [tests](#tests) section.

<table>
<col width='50px'><col width='300px'><col width='300px'>
<tr>
  <th>&nbsp;
  <th>Pros
  <th>Cons

<tr>
<th>stopifnot
<td>
  <ul>
  <li>No learning curve
  <li>Fast
  <li>Part of base
  </ul>
<td>
  <ul>
  <li>Comprehensive parameter checking is tedious
  <li>Error messages can seem cryptic
  </ul>
<tr>
<th>asserthat
<td>
  <ul>
  <li>No learning curve
  <li>Better error messages for basic `is.*` checks
  <li>A few helpful additional `is.*` functions for common checks
  </ul>
<td>
  <ul>
  <li>Comprehensive parameter checking is tedious
  </ul>
<tr>
<th>assertive
<td>
  <ul>
  <li>Custom check functions for most situations
  <li>Excellent error reporting
  <ul>
  <li>Reports what the object is in addition to what it is not
  <li>Will point out exact elements that fail and why
  </ul>
  </ul>
<td>
  <ul>
  <li>Slowest of the group, by far
  <li>Developer must wade through hundreds of functions to find the right ones,
  and in our simple suite of tests we could not find functions well suited to
  our requirements so we ended up with fairly verbose checks.
  </ul>
<tr>
<th>checkmate
<td>
  <ul>
  <li>Custom check functions for most situations
  <li>Fast, particularly for vectors
  <li>Improved error reporting
  </ul>
<td>
  <ul>
  <li>Dozens of reasonably complex functions
  </ul>
<tr>
<th>ensurer
<td>
  <ul>
  <li>Easily extensible checking framework
  <li>Easily customizable failure conditions
  <li>Can check function return values
  </ul>
<td>
  <ul>
  <li>Fairly slow
  <li>Introduces several new concepts with a corresponding learning curve
  </ul>
<tr>
<th>vetr
<td>
  <ul>
  <li>Intuitive and concise declarative syntax for checking objects structure
  via templates
  <li>Fast, particularly for complex objects and large vectors
  <li>Very extensible: vetting expressions are trivially re-usable language
  objects
  <li>Emphasis on useful error reporting, particularly for templates:
  <ul>
  <li>Reports what the object is in addition to what it is not (for templates)
  <li>Will point out exact elements that fail and why (for templates)
  <li>Displays the failing parameter as it appears in the calling context
  <li>Displays the failing call
  </ul>
<td>
  <ul>
  <li>Introduces new concepts (templates, like those in `vapply`)
  <li>Still in development, although interface should be mostly stable.
  </ul>
</table>

### Ease of Use for Function Developer

We assess this category based on how easy it is to learn how to use each
package, and by how concise and extensible those validation mechanisms they
implement are.

<table>
<tr>
  <th>&nbsp;
  <th colspan=2>Learning Curve
  <th colspan=2>Expressive Power
  <th colspan=2>Extensibility

<tr>
  <th>asserthat
  <td style='weight: bold;'>&nbsp;0<td>like `stopifnot`
  <td style='weight: bold;'>&nbsp;0<td>like `stopifnot`
  <td style='weight: bold;'>&nbsp;0<td>like `stopifnot`
<tr>
  <th>assertive
  <td style='color: red; weight: bold;'>-2<td>hundreds of functions
  <td style='color: green; weight: bold;'>+1<td>Functions for most circumstances
  <td style='weight: bold;'>&nbsp;0<td>like `stopifnot`
<tr>
  <th>checkmate
  <td style='color: red; weight: bold;'>-2<td>dozens of complex functions
  <td style='color: #0F0; weight: bold;'>+2<td>Excellent for vectors
  <td style='color: green; weight: bold;'>+1<td>Tools to build custom assertions
<tr>
  <th>ensurer
  <td style='color: orange; weight: bold;'>-1<td>new concepts
  <td style='weight: bold;'>&nbsp;0<td>like `stopifnot`
  <td style='color: green; weight: bold;'>+1<td>Tools to build custom assertions
<tr>
  <th>vetr
  <td style='color: orange; weight: bold;'>-1<td>new concepts
  <td style='color: #0F0; weight: bold;'>+2<td>Excellent for complex object
  structure
  <td style='color: #0F0; weight: bold;'>+2<td>Extensible vetting expressions
</table>

`assertthat` hews very closely to `stopifnot` so it basically doesn't have any
learning curve.  It does implement a few useful convenience check functions.
`assertive` and `checkmate` implement many check functions which increases the
expressive power, but also the learning curve.  Even when accounting for the
duplicate/variation functions in `assertive` and `checkmate`, the overall
function count is substantially higher than in the other packages (see
[appendix](#appendix) for `count_funs`):

```{r list_funs, echo=FALSE}
list_funs <- function(pkg) {
  space <- sprintf("package:%s", pkg)
  object.names <- ls(space)
  objects <- mget(object.names, envir=as.environment(space))
  names(objects)[sapply(objects, is.function)]
}
count_funs <- function(pkgs) {
  fun.list <- lapply(pkgs, list_funs)
  lens <- lengths(fun.list)
  names(lens) <- sprintf(
    "`%s (v%s)`", pkgs, as.character(lapply(pkgs, packageVersion))
  )
  lens
}
```
```{r}
count_funs(c('vetr', 'assertthat', 'ensurer', 'assertive', 'checkmate'))
```

Additionally, in the limited experience of putting together these comparisons
the intersection between the checks we wanted and the available functions was
surprisingly limited.  `checkmate` has more complex check functions which
increases the learning curve further, but allows for some very concise and
effective checks.

`ensurer` takes a different angle, focusing mostly on tools that help generate
custom checking functions (contracts) and mechanisms for attaching them to
functions.

`vetr` introduces the concept of templates, with an accompanying learning
curve.  The reward is the ability to express complex structural requirements
intuitively **and** succinctly.  `vetr` relies on standard R expressions like
`stopifnot` for non-structural checks.

### Performance

<table>
<tr>
  <th>&nbsp;
  <th colspan=2>asserthat <th colspan=2>assertive <th colspan=2>checkmate
  <th colspan=2>ensurer <th colspan=2>vetr

<tr>
<th>Simple checks
  <td style='weight: bold;'>&nbsp;0<td>comparable to `stopifnot`
  <td style='color: red; weight: bold;'>-2<td>2 orders of magnitude slower
  <td style='weight: bold;'>&nbsp;0<td>comparable to `stopifnot`
  <td style='color: orange; weight: bold;'>-1<td>1 order of magnitude slower
  <td style='weight: bold;'>&nbsp;0<td>comparable to `stopifnot`

<tr>
<th>Large vectors
  <td style='weight: bold;'>&nbsp;0<td>comparable to `stopifnot`
  <td style='color: red; weight: bold;'>-2<td>3 orders of magnitude slower
  <td style='weight: bold; color: green;'>+1<td>Comparisons in C
  <td style='weight: bold;'>&nbsp;0<td>comparable to `stopifnot`
  <td style='weight: bold; color: green;'>+1<td>Comparisons in C
<tr>

<th>Complex Objects
  <td style='weight: bold;'>&nbsp;0<td>comparable to `stopifnot`
  <td style='color: red; weight: bold;'>-2<td>3 orders of magnitude slower
  <td style='weight: bold; color: green;'>+1<td>Comparisons in C
  <td style='weight: bold;'>&nbsp;0<td>comparable to `stopifnot`
  <td style='weight: bold; color: green;'>+1<td>Comparisons in C
</table>

### User Friendliness

<table>
<tr>
  <th>&nbsp;
  <th colspan=2>asserthat <th colspan=2>assertive <th colspan=2>checkmate
  <th colspan=2>ensurer <th colspan=2>vetr

<tr>
<th>Explains failure
  <td style='weight: bold; color: green;'>+1<td>Improved messages for
  base `is.*` functions
  <td style='color: #0F0; weight: bold;'>+3<td>Improved messages for all
  defined check functions, including specific details on failure.
  <td style='color: #0F0; weight: bold;'>+2<td>Improved messages for all
  defined check functions, some provide details on failure.
  <td style='color: #0F0; weight: bold;'>+2<td>Improved messages for all
  defined check functions.
  <td style='color: #0F0; weight: bold;'>+2<td>Improved messages for template
  based checks, including specific details on failure (also for value checks
  with `all_bw`).
<tr>
<th>Parameter as it appears in calling frame
  <td style='weight: bold;'>&nbsp;0<td>No
  <td style='weight: bold;'>&nbsp;0<td>No
  <td style='weight: bold;'>&nbsp;0<td>No
  <td style='weight: bold;'>&nbsp;0<td>No
  <td style='color: #0F0; weight: bold;'>+2<td>Yes

<tr>
<th>Reports `sys.call`
  <td style='weight: bold;'>&nbsp;0<td>No
  <td style='weight: bold;'>&nbsp;0<td>No
  <td style='weight: bold;'>&nbsp;0<td>No, returns topmost call
  <td style='weight: bold;'>&nbsp;0<td>No, returns topmost call
  <td style='color: #0F0; weight: bold;'>+2<td>Yes
</table>

Ideally check functions would tell you as much as possible about the nature of a
failure.  For example, in addition to telling you what a parameter is not, it is
useful to know what it is, or if a value is invalid, where in that object that
value is.  This is one area where `assertive` shines.  `checkmate` and `vetr`
will also report additional information.

Context is also helpful.  `vetr` returns both the failing function call so you
can see what function caused the problem, as well as the parameter as input into
the calling frame.  The latter is helpful because it means you can copy paste
the expression from the error message into the calling environment to examine
the value that causes the parameter to fail.

## Old Stuff

## Simple Parameter Checks

### The Function

We want to check the following three parameter function:

```{r, eval=FALSE}
function(
  num,     # numeric vector
  opt,     # scalar character in c("one", "two", "three")
  flag     # TRUE or FALSE
)
```
```{r}
valid.opts <- c("one", "two", "three")   # values valid in `opts`
```

### Wrapper Based

We start with `stopifnot`, since it is the simplest and most well known option:

```{r}
f1_stopifnot <- function(num, opt, flag) {
  stopifnot(
    is.numeric(num),
    isTRUE(opt %in% valid.opts),
    length(flag) == 1 && !is.na(flag) && is.logical(flag)
  )
  TRUE
}
f1_stopifnot(letters, "four", TRUE)
```

The `vetr` implementation is very similar to `stopifnot`, except we can use
templates and `.` as shorthand for the matched parameter:

```{r}
f1_vetr <- function(num, opt, flag) {
  vetr(
    numeric(),                 # template
    isTRUE(. %in% valid.opts), # custom token (contains `.`)
    logical(1L) && !is.na(.)   # template && custom token
  )
  TRUE
}
f1_vetr(letters, "four", TRUE)
```

In fact, we *must* use `.` if we want `vetr` to evaluate a token as an R
expression instead of treating it as a template.  Notice how the template
`logical(1L)` is equivalent to `length(flag) == 1 && is.logical(flag)`.
Parameters must match the structure of their corresponding templates.  For more
details on templates and validation expressions see the vignette
`vignette('vetr', package='vetr')`.

Worth noting:

* Parameters are positionally matched to vetting expressions to save typing,
  although you can use names as well (e.g. `vetr(num=numeric(), ...)`)
* In addition to being told what the object isn't, you are also told what it is
* The object is referenced by its name in the calling scope (`letters`) instead
  of its name in the function environment (`num`)
* The function call is returned so you know where the error happened

Let's continue on with the other wrapper style packages:

```{r}
f1_assertthat <- function(num, opt, flag) {
  assert_that(
    is.numeric(num),
    opt %in% valid.opts,
    is.flag(flag) && !is.na(flag)
  )
  TRUE
}
f1_assertthat(letters, "four", TRUE)

f1_ensurer <- function(num, opt, flag) {
  ensure(num, is.numeric(.))
  ensure(opt, . %in% valid.opts)
  ensure(flag, length(.) == 1 && !is.na(.) && is.logical(.))
  TRUE
}
f1_ensurer(letters, "four", TRUE)

f1_valaddin <- firmly(
  function(num, opt, flag) TRUE,
  list(~num) ~ is.numeric, # equivalently `vld_numeric(~num)`
  list(~opt) ~ {. %in% valid.opts},
  list(~flag) ~ {length(.) == 1 && !is.na(.) && is.logical(.)}
)
f1_valaddin(letters, "four", TRUE)
```

Worth noting:

* `asserthat`, `valaddin`, and `ensurer` only pass if test expressions return
  single TRUE values, unlike `stopifnot` and `vetr` which also accept all TRUE
  logical vectors
* Wrapper style packages can implement specific checker functions: for example
  `assertthat` implements `is.flag` function, `valaddin` implements
  `vld_boolean` (unused here), etc.
* `valaddin`:
    * reports all failing conditions, not just the first failing one
    * reports the failing call, just like `vetr`
    * modifies an *existing* function

### Function Based

```{r}
f1_checkmate <- function(num, opt, flag) {
  assertNumeric(num)
  assertChoice(opt, valid.opts)
  assertFlag(flag)
  TRUE
}
f1_checkmate(letters, "four", TRUE)

f1_assertive <- function(num, opt, flag) {
  assert_is_numeric(num)
  assert_is_scalar(opt)
  assert_all_are_true(opt %in% valid.opts)
  assert_is_a_bool(flag)
  assert_all_are_not_na(flag)
  TRUE
}
f1_assertive(letters, "four", TRUE)
```

The "function" based packages have specialized functions for a variety of
checks.  Sometimes this can lead to particularly concise checks, and also more
helpful error messages since each specific function has more context about the
validation expectations.

### Benchmarks

We benchmark the functions with `mb`, a thin wrapper around `microbenchmark`
(see the [appendix](#appendix) for its definition).  Note that the `mean`
numbers are likely skewed by outlier timings caused by `rmarkdown`.

```{r mb_define, echo=FALSE}
mb <- function(...) {
  mb.call <- match.call()
  mb.call[[1]] <- quote(microbenchmark::microbenchmark)

  mb.res <- summary(eval(mb.call, envir=parent.frame()))
  mb.res <- mb.res[order(mb.res$median), ]
  mb.res[, -1] <- lapply(
    mb.res[, -1], function(x) format(round(x), big.mark=",")
  )
  cat(
    sprintf("Unit: %s, neval: %s\n\n", attr(mb.res, 'unit'), mb.res$neval[1])
  )
  print(mb.res[, c('expr', 'lq', 'median', 'uq', 'mean')], quote=FALSE)
}
```
```{r, echo=FALSE}
mb(
  f1_assertive(pi, "three", TRUE),
  f1_assertthat(pi, "three", TRUE),
  f1_ensurer(pi, "three", TRUE),
  f1_checkmate(pi, "three", TRUE),
  f1_stopifnot(pi, "three", TRUE),
  f1_valaddin(pi, "three", TRUE),
  f1_vetr(pi, "three", TRUE)
)
```

`stopifnot` leads the way, which is not surprising given the simplicity of the
checks involved.  `checkmate` is second, although both `vetr` and `asserthat`
are in the same neighborhood.  Performance drops off substantially after
`asserthat`.

One could argue that the performance differences here are trivial.  After all,
even the slowest approach clocks in under 3 milliseconds.  However, given the
prevalence of split-apply-combine style analysis, where a single function could
be applied to thousands or even millions of data groups, the performance
differences shown here can have a substantial impact on overall computation
time.

## Strictly Positive

We will increase the complexity somewhat by requiring that the numeric vector be
strictly positive.

```{r}
f2_vetr <- function(num) {
  vetr(numeric() && all_bw(., 0, Inf,  bounds="(]"))
}
f2_stopifnot <- function(num) {
  stopifnot(is.numeric(num), !anyNA(num), all(num > 0))
}
f2_assertthat <- function(num) {
  assert_that(is.numeric(num), !anyNA(num), all(num > 0))
}
f2_ensurer <- function(num) {
  ensure(num, is.numeric(.), !anyNA(num), all(num > 0))
}
f2_valaddin <- firmly(
  function(num) NULL,
  ~ is.numeric, ~ {!anyNA(.)}, ~ {all(. > 0)}
)
f2_checkmate <- function(num) {
  qassert(num, "n*(0,]")
}
f2_assertive <- function(num) {
  assert_all_are_in_left_open_range(num, lower=0)
}
```
```{r}
str.pos.vec <- runif(5e4) + 1  # test with a 50K long vector
```
```{r, echo=FALSE}
mb(times=10,
  f2_assertive(str.pos.vec),
  f2_assertthat(str.pos.vec),
  f2_ensurer(str.pos.vec),
  f2_checkmate(str.pos.vec),
  f2_stopifnot(str.pos.vec),
  f2_valaddin(str.pos.vec),
  f2_vetr(str.pos.vec)
)
```

`vetr` is now the fastest option, although this is primarily because we use
`vetr::all_bw` instead of the equivalent `!anyNA(.) && all(. > 0)` expression.
`all_bw` is implemented in C and avoids the intermediate vectors required to
evaluate the standard R version.  `checkmate` does the same with `qassert`.
Notice also that for both `vetr` and `checkmate` the mean and median timings are
close.  This is likely because these functions do not require the
occasional expensive garbage collections caused by intermediate vectors.

## A Complex Object

Here we wish to verify that an input conforms to the structure of the `iris`
built-in data set.

```{r}
iris.tpl <- vetr::abstract(iris) # equivalently: iris[0, ]
iris.fake <- iris
levels(iris.fake$Species)[3] <- "sibirica"   # tweak levels
```
We will narrow the field to `vetr`, `stopifnot`, and `checkmate` primarily
```{r}
f3_vetr <- function(x) {
  vetr(iris.tpl)
}
f3_vetr(iris.fake)

f3_stopifnot <- function(x) {
  stopifnot(
    is.list(x), inherits(x, "data.frame"),
    length(x) == 5, is.integer(attr(x, 'row.names')),
    identical(names(x), names(iris.tpl)),
    typeof(x$Species) == "integer", is.factor(x$Species),
    identical(levels(x$Species), levels(iris.tpl$Species))
  )
}
f3_stopifnot(iris.fake)

f3_checkmate <- function(x) {
  assertDataFrame(
    x, types=c("numeric", "numeric", "numeric", "numeric", "factor")
  )
  assertNames(names(x), identical.to=names(iris.tpl))
  assertFactor(x$Species, levels=levels(iris.tpl$Species))
}
f3_checkmate(iris.fake)
```

This is where the power of the template becomes apparent.  Rather than
painstakingly generating a list of all the comparisons required to validate the
object, we point to an existing object to use as a template. `vetr` recursively
traverses the template and the function parameter in parallel, ensuring that the
structure of the template elements is matched by those of the parameter
elements.  Even if an existing object does not exist, it is easier to manually
construct such a template than to specify all the function calls required to
confirm conformance to the template structure.

In addition to the compact representation, notice the error message:

> `levels(iris.fake$Species)[3]` should be "virginica" (is "sibirica")

This is by far the most useful of the three as it highlights the exact location
of the failure in a form you can immediately copy-paste into your prompt for
further examination.

Additionally, since all the `vetr` comparisons are done internally in C, `vetr`
is the fastest option:

```{r, echo=FALSE}
mb(
  f3_vetr(iris),
  f3_stopifnot(iris),
  f3_checkmate(iris)
)
```

## Comparison Dimensions

* Error message
  * Reports call like stop
  * Explains what it is not
  * Uses language from calling scope
  * Explains what it is
* Style (wrapper / function)
* Median timings
    * Simple tests
    * Values
    * Data frame
    * S3 Object
* Compound expressions

### Or Checks

## Comparison

## Appendix

```{r list_funs}
```

```{r mb_define}
```

