---
title: "Parameter Validation Functions"
author: "Brodie Gaslam"
output:
    rmarkdown::html_vignette:
        toc: true
        css: ../vignettes/styles.css

vignette: >
  %\VignetteIndexEntry{vetr}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(error=TRUE, comment=NA)
library(vetr)
library(assertive)
library(assertthat)
library(ensurer)
library(checkmate)
library(valaddin)
```

## Prevent Garbage In

Most functions expect their parameters to be of specific types, structures, or
values.  If we allow functions to operate on parameters that do not conform to
the expectations then we could get in the best case difficult to debug errors
from the bowels of the function, and in the worst silent undefined behavior.

Systematically vetting function parameters is as tedious as it is important, but
fortunately many packages exist to assist with the process.  We will review
those packages and assess them across these dimensions:

* Ease of use for the function developer incorporating parameter checks
* Computing overhead
* Ease of use for end users

We will review the following packages / functions:

* `stopifnot` by R Core, evaluates test expressions and `stops` if they return
  anything other than all TRUE vectors.
* [`vetr`](https://github.com/brodieG/vetr) by Yours Truly, augments test
  expressions with templates for object structure verification.  This allows for
  a succinct and extensible declarative syntax.  The package also emphasizes
  informative error messages and speed.
* [`asserthat`](https://github.com/hadley/assertthat) by Hadley Wickham, a
  drop-in replacement for `stopifnot` with more helpful error messages for
  checks based on the base `is.*` functions (e.g. `is.character`) and some
  additional helper functions for common checks.
* [`assertive`](https://www.r-pkg.org/pkg/assertive) by Richie Cotton, a
  large library of checking functions for most common parameter checks with
  expressive error messages.
* [`checkmate`](https://github.com/mllg/checkmate) by Michael Lang, a library of
  flexible checking functions for most common types of objects, with an emphasis
  on speed and conciseness, particularly for vectors.
* [`ensurer`](https://github.com/smbache/ensurer) by Stefan M. Bache, a
  framework for constructing re-usable and extensible check contracts, with a
  flexible reporting channel, a mechanism for applying check contracts to
  function _outputs_, and an experimental interface for embedding the checking
  contracts into functions.

Not all of these packages are specifically designed to test function parameters,
but they can all be used that way.  We will also focus on "out of the box"
functionality.  For example, most of these packages support custom user supplied
error messages, but we evaluate only the error messages produced without
additional user intervention.  We also ignore functionality that is not
specifically used for function parameter checking.

The following packages also perform related tasks, although we do not review
them here:

* [`valaddin`](https://github.com/egnha/valaddin) by Eugene Ha, a framework for
  augmenting existing functions with validation contracts.  Currently the
  package is undergoing a major overhaul so we will add it to the comparison
  once the new release (v0.3.0) is out.
* [`validate`](https://github.com/data-cleaning/validate) by Mark van
  der Loo and Edwin de Jonge, with a primary focus on validating data in data
  frames and similar data structures
* [`assertr`](https://github.com/ropensci/assertr) by Tony Fischetti, also
  focused on data validation in data frames and similar structures
* [`types`](https://github.com/jimhester/types) by Jim Hester, which implements
  but does not enforce type hinting
* [`argufy`](https://github.com/gaborcsardi/argufy) by Gábor Csárdi, which
  implements parameter validation via roxygen tags (not released to CRAN)

## Summary

Summary of philosophical objectives of each package:

### Simplicity

`assertthat` hews very closely to `stopifnot` semantics so it is trivial to
learn<sup><a href='#fn-1'>1</a></sup>.

Every other package adds either function libraries, or new concepts that need to
be mastered:

* `assertive`, `checkmate`: extensive libraries of Predicate Functions (e.g.
  `is.something(...)`).  `assertive` focuses on many simple single argument
  Predicate Functions, whereas `checkmate` has a smaller library of more
  flexible and complex multi-argument Predicate Functions.
* `vetr`: declarative template based checks for object structure.
* `ensurer`: convert Predicate Expressions into contracts.

### Expressive Power

The trade-off to increased complexity is expressive power, which allows the user
to express complex requirements succinctly:

* `checkmate`: provides highly configurable functions, as well as a domain
  specific language for vector checks.
* `vetr`: verifies structural requirements by comparison to template objects.

`assertive` is focused on single parameter Predicate Functions, which by nature
are limited in their expressive power.  `ensurer` and `assertthat` rely on
standard R expressions so are equivalent to `stopifnot` in this respect.

### Informative Failure Messages

* `assertthat`: clearer error messages for the base `is.*`, `any`, `all`, and
  the handful of bundled Predicate Functions, although no additional information
  is provided except for `any` and `all` (position of failure).
* `assertive`, and `checkmate`: clearer error messages for their Predicate
  Functions, including additional information, particularly for `assertive`.
* `vetr`: clearer error messages with additional informations for object
  structure (type, length, attributes), and for values with `all_bw`.

All packages allow users to attach custom error messages, but here we focus on
those available "out of the box".

### Speed

* `stopifnot` is as fast as the R expressions you use for it.
* `assertthat` is comparable but adds some overhead.
* `checkmate` and `vetr` are slower for trivial checks, but faster for large
  vectors, and for `vetr`, complex objects.
* `ensurer` is slower
* `assertthat` is substantially slower

<table>
<col width='2*'><col width='*'><col width='*'><col width='*'><col width='*'>
<thead>
<tr>
  <th><th>Simple<th>Expressive<th>Informative<th>Fast
<tbody>
<tr>
  <td>`assertthat` <td style='color: green;'>*✔️*<td><td><td>
<tr>
  <td>`assertive`  <td><td><td><td>
<tr>
  <td>`checkmate`  <td><td><td><td>
<tr>
  <td>`ensurer`    <td><td><td><td>
<tr>
  <td>`vetr`       <td><td><td><td>
</table>


<table>
<tr>
  <th>&nbsp;
  <th>Pros
  <th>Cons

<tr>
<th>stopifnot
<td>
  <ul>
  <li>No learning curve
  <li>Fast
  <li>Part of base
  </ul>
<td>
  <ul>
  <li>Comprehensive parameter checking is tedious
  <li>Error messages can seem cryptic
  </ul>
<tr>
<th>asserthat
<td>
  <ul>
  <li>No learning curve
  <li>Better error messages for basic `is.*` checks
  <li>A few helpful additional `is.*` functions for common checks
  </ul>
<td>
  <ul>
  <li>Comprehensive parameter checking is tedious
  <li>Fast, but slower than `stopifnot`
  </ul>
<tr>
<th>assertive
<td>
  <ul>
  <li>Custom check functions for most situations
  <li>Excellent error reporting
  </ul>
<td>
  <ul>
  <li>Slowest of the group, by far
  <li>Hundreds of functions
<tr>
<th>checkmate
<td>
  <ul>
  <li>Custom check functions for most situations
  <li>Fast, particularly for vectors
  <li>Improved error reporting
  </ul>
<td>
  <ul>
  <li>Dozens of reasonably complex functions
  </ul>
<tr>
<th>ensurer
<td>
  <ul>
  <li>Easily extensible checking framework
  <li>Easily customizable failure conditions
  </ul>
<td>
  <ul>
  <li>Fairly slow
  <li>New concepts
  </ul>
<tr>
<th>vetr
<td>
  <ul>
  <li>Intuitive and concise declarative object structure checks
  <li>Fast, particularly for complex objects and large vectors
  <li>Trivially extensible vetting expressions
  <li>Emphasis on useful error reporting
  </ul>
<td>
  <ul>
  <li>New concepts
  <li>Still in development, although interface should be mostly stable.
  </ul>
</table>

`asserthat` is the package that makes fewest trade-offs relative to `stopifnot`.
However the improvements it makes are modest.

`assertive` and `checkmate` improve on `stopifnot` by packaging a
broad range of common checks into a function library (a very large library for
`assertive`, or a modest one with complex functions for `checkmate`).  This adds
a learning curve, and in the case of `assertive` a lot of computation overhead.

`ensurer` and `vetr` improve on `stopifnot` by introducing new validation
concepts.  This also adds to the learning curve, and in the case of `ensurer`
adds a bit of computation overhead.

```{r list_funs, echo=FALSE}
list_funs <- function(pkg) {
  space <- sprintf("package:%s", pkg)
  object.names <- ls(space)
  objects <- mget(object.names, envir=as.environment(space))
  names(objects)[sapply(objects, is.function)]
}
count_funs <- function(pkgs) {
  fun.list <- lapply(pkgs, list_funs)
  lens <- lengths(fun.list)
  names(lens) <- sprintf(
    "`%s (v%s)`", pkgs, as.character(lapply(pkgs, packageVersion))
  )
  lens
}
```
```{r}
count_funs(c('vetr', 'assertthat', 'ensurer', 'assertive', 'checkmate'))
```

## Comparisons

### Simple Parameters

We want to write a function that accepts a numeric vector as the first
argument, and TRUE or FALSE as the second.  We start with `stopifnot`, since it
is the simplest and most well known option:

```{r}
f1_stopifnot <- function(num, flag) {
  stopifnot(
    is.numeric(num),
    is.logical(flag), length(flag) == 1, !is.na(flag)
  )
}
```

`vetr` looks similar on the surface, but it matches each of its parameters
to those of the enclosing function we:

* use `.` instead of the parameter name
* combine tokens with `&&` (or `||`) instead of commas

```{r}
f1_vetr_a <- function(num, flag) {
  vetr(
    is.numeric(.),                                # matched to `num`
    is.logical(.) && length(.) == 1 && !is.na(.)  # matched to `flag`
  )
}
```

Additionally, `vetr` introduces the concept of templates which allow you to
specify object structure declaratively as in `vapply`, so we can rewrite
`f1_vetr_a` as:

```{r}
f1_vetr_b <- function(num, flag) {
  vetr(
    numeric(),                 # template
    logical(1L) && !is.na(.)   # template && standard token
  )
}
```

`flag` must be type `logical` and length 1 to match `logical(1L)`.  Similarly
`num` must be numeric to match `numeric()` (we allow length zero
dimensions in templates to match any length dimension in the parameter).  `vetr`
distinguishes between template tokens and standard tokens by the absence or
presence of the `.` symbol.

`asserthat` is almost identical to `stopifnot`, with the stricter
requirement that its arguments must evaluate to TRUE or FALSE exactly, and
the addition of helper Predicate Functions like `is.flag`:

```{r}
f1_assertthat <- function(num, flag) {
  assert_that(
    is.numeric(num),
    is.flag(flag), !is.na(flag)
  )
}
```

`ensurer` is also similar approach, although requires a separate call to
`ensure` for each argument:

```{r}
f1_ensurer <- function(num, flag) {
  ensure(num, is.numeric(.))
  ensure(flag, length(.) == 1 && !is.na(.) && is.logical(.))
}
```

`assertive` implements 200+ Predicate Functions and assertion based on them
instead of relying on standard R expressions:

```{r}
f1_assertive <- function(num, flag) {
  assert_is_numeric(num)
  assert_is_a_bool(flag)
  assert_all_are_not_na(flag)
}
```

`checkmate` implements 40+ Predicate Functions and assertions based on them, but
they are more configurable (and complex) than those in `assertive`:

```{r}
f1_checkmate <- function(num, flag) {
  assertNumeric(num)
  assertFlag(flag)
}
```

For this type of check it is not obvious that any of the packages do a much
better job than `stopifnot` from the perspective of the developer.  There are
however some noticeable differences when we actually run the functions with
illegal parameters:

```{r}
f1_stopifnot(letters, TRUE)
f1_vetr_b(letters, TRUE)
```

In addition to what our parameter is not, `vetr` tells you what it is, gives you
the original call of the function, and gives you the expression that corresponds
to the parameter (`letters`) instead of just the parameter name.  This
additional contextual information makes debugging easier.

The other packages include some of this additional information:

```{r}
f1_assertthat(letters, TRUE)
f1_assertive(letters, TRUE)
f1_ensurer(letters, TRUE)
f1_checkmate(letters, TRUE)
```

`assertive`, and `checkmate` tell you what the object is.  `ensurer` and
`checkmate` return the top level call, although that does not match the
semantics of a `stop` call in a function, and is in our opinion less useful.

### Building Compound Checks

If you wish to build re-usable complex checks with `stopifnot`  or `asserthat`
you would do so by writing functions.  We illustrate by creating a custom check
for either a square numeric matrix or a scalar numeric:

```{r}
is.sqr_mx <- function(x) is.matrix(x) && ncol(x) == nrow(x)
is.sqr_num_mx_or_sclr_num <- function(x)
  is.numeric(x) && (is.sqr_mx(x) || length(x) == 1)

f2_stopifnot <- function(x) {
  stopifnot(is.sqr_num_mx_or_sclr_num(x))
}

rect.mx <- matrix(1:12, 3)
f2_stopifnot(rect.mx)
```

The main drawback from this approach is the somewhat uninformative error
message.

`vetr` is similar, except that you can combine template and standard tokens as
language objects and retain the more granular error reporting:

```{r}
vetr.sqr.num.mx <- quote(matrix(numeric(), 0) && ncol(.) == nrow(.))
vetr.sqr.num.mx.or.sclr.num <- quote(vetr.sqr.num.mx || numeric(1L))

f2_vetr <- function(x) {
  vetr(vetr.sqr.num.mx.or.sclr.num)
}
f2_vetr(rect.mx)
```

You can use `assert` with `check*` functions in, but we lose some of the
expressiveness of the error messages:

```{r}
checkSquareNumMatrix <- function(x)
  isTRUE(checkMatrix(x, mode='numeric')) && nrow(x) == ncol(x)

f2_checkmate <- function(x) {
  assert(checkSquareNumMatrix(x), checkNumber(x, na.ok=TRUE), combine="or")
}
f2_checkmate(rect.mx)
```

Neither `assertive` nor `ensurer` appear to provide an out of the box mechanism
for combining checks with "or" logic.  `ensurer` does provide a mechanism for
"and" combinations [via `ensures`](#appendix).  To get the "or" logic we need to
rely on the same function we defined for `stopifnot`:

```{r}
f2_ensurer <- function(x) {
  ensure(is.sqr_num_mx_or_sclr_num(x))
}
f2_ensurer(rect.mx)

f2_assertive <- function(x) {
  assert_engine(is.sqr_num_mx_or_sclr_num, x=x)
}
f2_assertive(rect.mx)
```

Outside of `stopifnot` every package has a mechanism for manually specifying
more informative error messages, but we do not explore those here.

### Vector Values

Suppose we wish to ensure our input is a strictly positive numeric vector with
no missing values.  With `stopifnot` we would
use:

```{r}
vec <- -1:1

f3_stopifnot <- function(num) {
  stopifnot(is.numeric(num), !anyNA(num), all(num > 0))
}
f3_stopifnot(vec)
```

`vetr` includes the `all_bw` function primarily for speed, but it also generates
more useful error messages:

```{r}
f3_vetr <- function(num) {
  vetr(numeric() && all_bw(., 0, Inf,  bounds="(]"))
}
f3_vetr(vec)
```

`asserthat` and `ensurer` end up very similar to `stopifnot`, though
`assertthat` improves the error message:

```{r}
f3_assertthat <- function(num) {
  assert_that(is.numeric(num), !anyNA(num), all(num > 0))
}
f3_assertthat(vec)

f3_ensurer <- function(num) {
  ensure(num, is.numeric(.), !anyNA(num), all(num > 0))
}
f3_ensurer(vec)
```

`checkmate` implements a particularly succinct notation:

```{r}
f3_checkmate <- function(num) {
  qassert(num, "n*(0,]")
}
f3_checkmate(vec)
```

`assertive` has a custom function for the job, with a particularly helpful
error message:

```{r}}
f3_assertive <- function(num) {
  assert_all_are_in_left_open_range(num, lower=0)
}
f3_assertive(vec)
```

### Complex Objects

Here we wish to verify that an input conforms to the structure of the `iris`
built-in data set.

```{r}
# collect data for tests
iris.tpl <- vetr::abstract(iris) # equivalently: iris[0, ]
iris.col.classes <- lapply(iris, class)

# make a bad version of iris
iris.fake <- iris
levels(iris.fake$Species)[3] <- "sibirica"   # tweak levels
```

Then, with `stopifnot`:

```{r}
f4_stopifnot <- function(x) {
  stopifnot(
    is.data.frame(x),                     # this only checks class
    is.list(x),
    length(x) == length(iris.tpl),
    identical(lapply(x, class), iris.col.classes),
    is.integer(attr(x, 'row.names')),
    identical(names(x), names(iris.tpl)),
    identical(typeof(x$Species), "integer"),
    identical(levels(x$Species), levels(iris.tpl$Species))
  )
}
f4_stopifnot(iris.fake)
```

In contrast, `vetr` is a one-liner:

```{r}
f4_vetr <- function(x) {
  vetr(iris.tpl)
}
f4_vetr(iris.fake)
```

Rather than painstakingly generating a list of all the comparisons required to
validate the object, we point to an existing object to use as a template. `vetr`
recursively traverses the template and the function parameter in parallel and
checks each sub element of the latter against the former.

`vetr` derives contextual information from the template that it uses to generate
the error messages:

```
`levels(iris.fake$Species)[3]` should be "virginica" (is "sibirica")
```

It is designed so you can copy all of or part of `levels(iris.fake$Species)[3]`
into the R prompt for examination.  It also tells you exactly what is wrong.

`checkmate` is reasonably succinct:

```{r}
f4_checkmate <- function(x) {
  assertDataFrame(x, types=unlist(iris.col.classes), ncols=5)
  assertTRUE(is.list(x))
  assertInteger(attr(x, 'row.names'))
  assertNames(names(x), identical.to=names(iris.tpl))
  assertFactor(x$Species, levels=levels(iris.tpl$Species))
}
f4_checkmate(iris.fake)
```

`assertive`, `assertthat`, and `ensurer`  end up with the same number of
explicit checks as `stopifnot`, and with similar error messages.  See the code
appendix for those implementations.

```{r iris.other, echo=F, eval=F}
f4_assertive <- function(x) {
  assert_is_list(x)
  assert_all_are_equal_to(length(iris), 5)
  assert_is_integer(attr(x, 'row.names'))
  assert_is_data.frame(x)
  assert_is_identical_to_true(identical(iris.col.classes, lapply(x, class)))
  assert_is_identical_to_true(identical(names(x), names(iris.tpl)))
  assert_is_identical_to_true(identical(typeof(x$Species), "integer"))
  assert_is_factor(x$Species)
  assert_is_identical_to_true(
    identical(levels(x$Species), levels(iris.tpl$Species))
  )
}
f4_assertive(iris.fake)

f4_assertthat <- function(x) {
  assert_that(
    is.data.frame(x),                     # this only checks class
    is.list(x),
    length(x) == length(iris.tpl),
    identical(lapply(x, class), iris.col.classes),
    is.integer(attr(x, 'row.names')),
    identical(names(x), names(iris.tpl)),
    identical(typeof(x$Species), "integer"),
    identical(levels(x$Species), levels(iris.tpl$Species))
  )
}
f4_assertthat(iris.fake)

f4_ensurer <- function(x) {
  ensure(
    x,
    is.data.frame(.),                     # this only checks class
    is.list(.),
    length(.) == length(iris.tpl),
    identical(lapply(., class), iris.col.classes),
    is.integer(attr(., 'row.names')),
    identical(names(.), names(iris.tpl)),
    identical(typeof(.$Species), "integer"),
    identical(levels(.$Species), levels(iris.tpl$Species))
  )
}
f4_ensurer(iris.fake)
```

## Benchmarks

We benchmark the functions with `mb`, a thin wrapper around `microbenchmark`
(see the [appendix](#appendix) for its definition).  Note that the `mean`
numbers are likely skewed by outlier timings caused by `rmarkdown`.

```{r mb_define, echo=FALSE}
mb <- function(...) {
  mb.call <- match.call()
  mb.call[[1]] <- quote(microbenchmark::microbenchmark)

  mb.res <- summary(eval(mb.call, envir=parent.frame()))
  mb.res <- mb.res[order(mb.res$median), ]
  mb.res[, -1] <- lapply(
    mb.res[, -1], function(x) format(round(x), big.mark=",")
  )
  cat(
    sprintf("Unit: %s, neval: %s\n\n", attr(mb.res, 'unit'), mb.res$neval[1])
  )
  print(mb.res[, c('expr', 'lq', 'median', 'uq', 'mean')], quote=FALSE)
}
```
```{r, echo=FALSE}
mb(
  f1_assertive(pi, "three", TRUE),
  f1_assertthat(pi, "three", TRUE),
  f1_ensurer(pi, "three", TRUE),
  f1_checkmate(pi, "three", TRUE),
  f1_stopifnot(pi, "three", TRUE),
  f1_valaddin(pi, "three", TRUE),
  f1_vetr(pi, "three", TRUE)
)
```

`stopifnot` leads the way, which is not surprising given the simplicity of the
checks involved.  `checkmate` is second, although both `vetr` and `asserthat`
are in the same neighborhood.  Performance drops off substantially after
`asserthat`.

One could argue that the performance differences here are trivial.  After all,
even the slowest approach clocks in under 3 milliseconds.  However, given the
prevalence of split-apply-combine style analysis, where a single function could
be applied to thousands or even millions of data groups, the performance
differences shown here can have a substantial impact on overall computation
time.

## Strictly Positive

```{r}
str.pos.vec <- runif(5e4) + 1  # test with a 50K long vector
```
```{r, echo=FALSE}
mb(times=10,
  f3_assertive(str.pos.vec),
  f3_assertthat(str.pos.vec),
  f3_ensurer(str.pos.vec),
  f3_checkmate(str.pos.vec),
  f3_stopifnot(str.pos.vec),
  f3_valaddin(str.pos.vec),
  f3_vetr(str.pos.vec)
)
```

`vetr` is now the fastest option, although this is primarily because we use
`vetr::all_bw` instead of the equivalent `!anyNA(.) && all(. > 0)` expression.
`all_bw` is implemented in C and avoids the intermediate vectors required to
evaluate the standard R version.  `checkmate` does the same with `qassert`.
Notice also that for both `vetr` and `checkmate` the mean and median timings are
close.  This is likely because these functions do not require the
occasional expensive garbage collections caused by intermediate vectors.

## A Complex Object

Additionally, since all the `vetr` comparisons are done internally in C, `vetr`
is the fastest option:

```{r, echo=FALSE}
mb(
  f4_vetr(iris),
  f4_stopifnot(iris),
  f4_checkmate(iris)
)
```

## Comparison Dimensions

* Error message
  * Reports call like stop
  * Explains what it is not
  * Uses language from calling scope
  * Explains what it is
* Style (wrapper / function)
* Median timings
    * Simple tests
    * Values
    * Data frame
    * S3 Object
* Compound expressions

### Or Checks

## Comparison

## Appendices


### Definitions

Definitions of terms as we use them in this document.  They may have different
definitions elsewhere.

<dl>
  <dt><a name='predicate-function'></a>Predicate Function</dt>
  <dd>
  Function that typically accepts one non-optional argument and returns TRUE
  or FALSE according to whether that argument conforms to a requirement.
  `isTRUE(x)`, `is.numeric(x)`, etc. are common examples of Predicate Functions.
  `identical(x, y)` is a two argument Predicate Function.
  </dd>
  <dt><a name='check-function'></a>Check Function</dt>
  <dd>
  Similar to a Predicate Function, arguments and may return either TRUE or a
  vector of only TRUE values on success, and something else on failure.
  `all.equal(x, y)` is a common example.  Check Functions should be used within
  `isTRUE(all(check_fun(...)))` to establish success or failure, unless they are
  used within `stopifnot` or `vetr` where such a check is implicit.
  </dd>
  <dt><a name='check-expression'></a>Check Expression</dt>
  <dd>
  A collection of calls to Check and/or Predicate Functions combined with
  logical operators.  For example `!anyNA(x) && x > 0`.
  </dd>
  <dt><a name='vetting-expression'></a>Vetting Expression</dt>
  <dd>
  Like a Check Expression, but augmented for use in `vetr`.  In addition to
  Check and Predicate Function calls, it can include Templates Tokens.
  References to the object being checked should be made with `.`.  For example
  `numeric(1L) && !is.na(.) && x > 0`, is made up of three tokens, where
  `numeric(1L)` is a Template Token, and `!is.na(.)` and `x > 0` are Standard
  Tokens.  See `vignette(package='vetr', 'vetr')` for more details..
  </dd>
  <dt><a name='template-token'></a>Template Token</dt>
  <dd>
  Part of a Vetting Expression that should be treated as a template by `vetr`.
  It should resolve to an R object when it is evaluated in the calling frame of
  the enclosing function to the `vetr` call.
  See `vignette(package='vetr', 'vetr')` for more details.
  </dd>
  <dt><a name='standard-token'></a>Standard Token</dt>
  <dd>
  Part of a Vetting Expression that should be treated as a standard R
  expression.  If `isTRUE(all(evaluated_standard_token))` then the token will be
  considered to pass.  Standard Tokens are identified by the presence of a `.`
  symbol within a token.  You can also wrap expressions in `.()` to mark them as
  standard tokens.  If you need to use the `.` symbol for other purposes escape
  it with another `.` (i.e. to use a literal `.`, use `..`).
  See `vignette(package='vetr', 'vetr')` for more details.
  </dd>
  <dt><a name='vetting-token'></a>Vetting Token</dt>
  <dd>
  Component pieces of Vetting Expressions.  May be either Standard Tokens or
  Template Tokens.  Vetting Tokens are the arguments to "top level" `&&`, `||`,
  and `(` calls.  For example, in the Vetting Expression
  `!anyNA(.) && I(. > 0 && interactive())` there are two  Vetting Tokens:
  `!anyNA(.)` and `I(. > 0 && interactive())`.  The `&&` inside `I(...)` is not
  considered to be at the "top level" because of the intervening `I` call.
  </dd>
</dl>

### References

* Vignettes and other documentation for the reviewed packages

### Ancillary Code

#### microbenchmark Thin Wrapper

```{r mb_define}
```

#### Ensurer Compound Contracts

`ensurer` "and" contracts:

```{r}
ensurer_num_mx <- ensures(is.numeric(.) && is.matrix(.))
ensurer_sqr <- ensures(ncol(.) == nrow(.))
ensurer_sqr_mx <- ensures(+ensurer_num_mx, +ensurer_sqr)  # and combination

f2_ensurer_a <- function(x) {
  ensurer_sqr_mx(x)
}
f2_ensurer_a(rect.mx)
```

#### Complex Objects Appendix

Complex objects for other packages:

```{r iris.other}
```

----
<sup>1</sup>:The only noteworthy difference between the two is that `assertthat`
requires Check Expressions to evaluate to TRUE or FALSE, whereas `stopifnot`
accepts anything, with all TRUE vectors considered a success and all else
failure.
<sup>1</sup>: Both `ensurer` and `checkmate` use the square matrix test in their
respective vignettes.


