---
title: "vetR - Trust, but Verify"
output:
    rmarkdown::html_vignette:
        toc: true
        css: vignettes/styles.css
---

```{r, echo=FALSE}
knitr::opts_chunk$set(error=TRUE)
library(vetr)
```

# Tweets

## Tweet #2

Fast #rstats bounds checks with `vetr::all_bw` https://github.com/brodieG/vetr;
10x improvement over primitives (1/3):

~30% of speedup is avoiding 3x1e6 temp logical vecs, rest is dedicated loops for
input permutations (num/int/char, nas, infs, etc.) (2/3).

Peeling back abstraction layer is eye-opening.  Massive complexity to implement something seemingly so simple in optimized manner (3/3).


~1/3 of speedup is avoiding 3
logical(1e6) intermediate vectors, rest is stripping inner loops to bare
minimum:


## Tweet #1

Declarative structural guarantees for #rstats S3 objects via templates with new
pkg #vetr https://github.com/brodieG/vetr (1/4)

### Like vapply

Declarative checks via templates, much like `vapply`:

```{r}
vet(numeric(1L), 1:3)
vet(numeric(1L), "hello")
vet(numeric(1L), 42)
```

### Handles Complex Objects

```{r}
vet(matrix(integer(), ncol=3), matrix(1:12, 4))
vet(matrix(integer(), ncol=3), matrix(1:12, 3))
```

Even recursive ones:

```{r}
iris.template <- abstract(iris)
levels(iris$Species)[3] <- 'Sibirica' # corrupt iris
```
```{r}
vet(iris.template, iris[1:10,])
```

Note the useful error messages.

## Tweet #2

We made `vetr` fast to mitigate overhead concerns. There is a
dedicated mode for fun param vetting (2/4)

### Fast

Comparable in performance to `stopifnot` for simple checks, and faster for
complex ones:

```{r}
mx.3.col.num <- matrix(numeric(), ncol=3)
mx1 <- matrix(1:12, 4)

bench_mark(times=1e4,
  vet(mx.3.col.num, mx1),
  stopifnot(is.matrix(mx1), is.numeric(mx1), ncol(mx1) == 3)
)
```

### In Functions

`vetr()` streamlines function parameter vetting:

```{r}
fun <- function(x, y) {
  vetr(integer(), character(1L) || NULL)
}
fun(1, 'hello')
fun(1, NULL)
fun(1, 2)
```

## Tweet #3

Create complex vetting expressions with intuitively programmable
non-standard-evaluation (3/4)

### Programmable NSE

`vetr` implements programmable NSE via recursive substitution of language
objects.  This allows you to construct complex vetting expressions from simple
ones:

```{r}
a <- quote(integer() && . > 0)
b <- quote(logical(1L) && !is.na(.))
```
```{r}
vet(a || b, 1:3)
vet(a || b, -1)
```
```{r}
c <- quote(a || b)  # equivalently
```
```{r}
vet(c, -1)
```

## Tweet #4

On CRAN, 100% coverage with #covr https://github.com/jimhester/covr
and #unitizer https://github.com/brodieG/unitizer, but under dev so feedback
welcome (4/4)

# R Package Announce

Announcing New Package vetr on CRAN

`vetr` implements a declarative template-based approach to verify that objects
meet structural requirements, and auto-compose error messages when they do not.
This package is intended to simplify a more formal use of S3 objects.

The template concept is borrowed from `vapply`:

>> vet(numeric(1L), 1:3)
>  [1] "`1:3` should be length 1 (is 3)"
>> vet(numeric(1L), "hello")
>  [1] "`\"hello\"` should be type \"numeric\" (is \"character\")"
>> vet(numeric(1L), 42)
>  [1] TRUE

There is no limit on template complexity:

>> vet(matrix(integer(), ncol=3), matrix(1:12, 3))
>  [1] "`matrix(1:12, 3)` should have 3 columns (has 4)"
>
>> iris.template <- abstract(iris)
>> levels(iris$Species)[3] <- 'sibirica'
>> vet(iris.template, iris[1:10,])
>  [1] "`levels((iris[1:10, ])$Species)[3]` should be \"virginica\"
>  (is \"sibirica\")"

`vetr` implements programmable non-standard evaluation via recursive
substitution of language objects:

>> a <- quote(integer() && . > 0)
>> b <- quote(logical(1L) && !is.na(.))
>> c <- quote(a || b)
>> vet(c, -1)
>  [1] "At least one of these should pass:"
>  [2] "  - `-1 > 0` is not TRUE (FALSE)"
>  [3] "  - `-1` should be type \"logical\" (is \"double\")"

Performance is comparable to `stopifnot` for simple checks, and is faster for
complex (template based) ones.  There is a mode that further streamlines
parameter vetting in functions.

The package is still under development, but the features should be mostly
stable.  Feedback welcome (https://github.com/brodieG/vetr).

Best regards,


Brodie.
