---
title: "validate - Keep the Garbage Out"
author: "Brodie Gaslam"
output:
    rmarkdown::html_vignette:
        toc: true
        css: styles.css

vignette: >
  %\VignetteIndexEntry{validate}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r global_options}
knitr::opts_chunk$set(error=TRUE)
```

## Overview

This package provides functions that simplify testing R objects for structural requirements.  The primary objective of the package is to make function validation of complex objects easier, and particularly to allow us to enforce structural requirements for S3 objects.  For example, here we use `validate_args` to require that a function have a 3 column numeric matrix as the first argument and a scalar logical as the second:

```{r}
fun <- function(a, b) {
  validate_args(matrix(numeric(), ncol=3), logical(1L))
  TRUE
}
fun(1:5, TRUE)
fun(matrix(1:6, 2), "TRUE")
fun(matrix(1:6, 3), TRUE)
fun(matrix(1:6, 2), TRUE)
```

`validate_args` uses R objects as templates that must be matched by the argument that is under validation.

`validate` is a utility function that allows you to validate a user specified object instead of automatically matching to a function's arguments:

```{r}
my.obj <- 1.5
validate(matrix(numeric(), ncol=3), my.obj)
```

We will primarily use `validate` in this vignette because it is simplifies exposition.  Validation expressions are used identically in `validate` and `validate_args`.  Unlike with `validate_args`, `validate` can optionally return the error message instead of `stop`ing:

```{r}
validate(matrix(numeric(), ncol=3), my.obj, return.mode="text")
```

## Templates

`validate` and `validate_args` use templates to define structural requirements that we wish to enforce.  Templates are "partially specified" R objects.  We check the value we want to test against the aspects of our template that are defined, and ignore all others.  Consider:

```{r}
validate(numeric(), runif(10))
validate(numeric(), sample(letters))
```

Our template is a zero (undefined) length numeric vector, so any numeric vector will satisfy it, but non-numeric objects will fail validation.  This becomes more interesting with complex objects:

```{r}
validate(numeric(), matrix(runif(9), 3))
```

The matrix validates because it is numeric.  Enforcing more constraits is straightforward.  In this example we require numeric matrices with three rows:

```{r}
mx.3.row <- matrix(numeric(), nrow=3)
validate(mx.3.row, matrix(runif(6), nrow=3))
validate(mx.3.row, matrix(runif(9), nrow=3))
validate(mx.3.row, matrix(runif(6), nrow=2))
```

Notice how both the two and three column matrices that have three rows pass validation, but the matrix that only has two rows does not.

The primary objective of template based validation is to allow us to enforce structural requirements for S3 objects.  Suppose that we are trying to build a program to analyze car test drives on race tracks:

```{r}
test.run.tpl <- structure(
  list(car.id=integer(1L), data=data.frame(dist=numeric(), time=Sys.time()[0])),
  class="test.run"
)
```
Then:
```{r}
analyze <- function(x, ...) UseMethod("analyze")
analyze.test.run <- function(x, ...) {
  validate_args(test.run.tpl)
  TRUE # stand-in for really meaningful code
}
```
Would allow us to ensure that only objects that actually match our test run template are fed through the `analyze` logic:

```{r}
run.1 <- structure(
  list(vehicle="mazda", data=data.frame(x=rep(TRUE, 10), y=runif(10))),
  class="test.run"
)
analyze(run.1)
names(run.1)[[1]] <- "car.id"
analyze(run.1)
run.1$car.id <- 548L
analyze(run.1)
run.1$car <- structure(list(model="mazda"), class="car")
analyze(run.1)
run.1$car <- structure(list(model="mazda", HP=250), class="car")
analyze(run.1)
run.1$data$time <- Sys.time() + 1:10
analyze(run.1)
run.1$data$x <- 1:10
analyze(run.1)
```

Internally `validate` uses `alike` to compare the template to the test value.  See the [alike vignette](http://htmlpreview.github.io/?https://raw.githubusercontent.com/brodieG/alike/master/inst/doc/alike.html) for extensive details on how templates work.

`alike` and `validate` both follow the same convention as `all.equal`, with the first argument the reference object to match (`target`) and the second the value to test (`current`).

## Validating Functions

`validate_args` is designed specifically to simplify testing function arguments.  For example, the following function requires a three column numeric matrix as an input, as well as an optional scalar logical argument:
```{r}
dist3d <- function(x, summarize=FALSE) {
  validate_args(matrix(numeric(), ncol=3), logical(1L))
  # do something
}
dist3d(matrix(runif(12), ncol=3))
dist3d(matrix(runif(12), ncol=3), summarize=TRUE)
dist3d(matrix(runif(12), ncol=4))
dist3d(matrix(runif(12), ncol=3), summarize=c(TRUE, FALSE))
```

Equivalently we could have written:

```{r}
dist3d <- function(x, summarize=FALSE) {
  stopifnot(
    is.numeric(x), is.matrix(x), ncol(x) == 3, is.logical(negate),
    length(negate) == 1L
  )
  # do something
}
dist3d(matrix(runif(12), ncol=3), summarize=TRUE)
dist3d(matrix(runif(12), ncol=3), summarize=c(TRUE, FALSE))
```

Both functions do the same validations, but with `validate_args` the validation is easier to specify, and the error messages are more helpul.

`validate_args` is very similar to `validate`, except you only have to specify the validation templates.  `validate_args` takes care of retrieving the values provided as the function arguments.  We could have written `dist3d` as:
```{r}
dist3d <- function(x, summarize=FALSE) {
  validate(matrix(numeric(), ncol=3), x)
  validate(logical(1L), summarize)
  # do something
}
```
AUTHNOTE: this will change if we change `validate` to return the reason for failure rather than `stop`.

## Complex Validation

**NOTE**: we will illustrate advanced features using `validate`, though you may use the same validation templates / expressions as arguments to `validate_args`.

### Multi-element Validation

Sometimes we want to allow an object to match different types of variables:

```{r}
validate(integer(1L) || character(1L), 2L)
validate(integer(1L) || character(1L), "hello")
validate(integer(1L) || character(1L), TRUE)
```

We are able to specify multiple templates by using `||`.

### Validation Expressions

At other times templates alone are not sufficient.  Here we check that values are length 3 numeric vectors and strictly positive:

```{r}
validate(numeric(3) && all(. > 0), c(1, 2, 3))
validate(numeric(3) && all(. > 0), c(1, -2, 3))
```

The `.` is a stand-in for the object being tested.

### Non Standard Evaluation of Validation Expressions

Validation expressions combine templates and custom expressions delimited by `&&` and `||`. While superficially validation expressions look like R expressions, they are never run through the R evaluator.  Instead, they are parsed into tokens and each token is examined to determine whether it should be treated as a template or as a validation expression.  For example:

```{r, eval=FALSE}
logical(1) || (numeric(1) && . %in% 0:1)
```

is parsed as three tokens: `logical(1)`, `numeric(1)`, and `. %in% 0:1`.  Tokens that contain `.` are interpreted as custom expressions, and tokens that don't are treated as templates.  In this case the first two tokens are considered templates, and the last is a custom expression, so the overall validation expression will accept scalar logicals, or 0 or 1:

```{r}
validate(logical(1) || (numeric(1) && . %in% 0:1), TRUE)
validate(logical(1) || (numeric(1) && . %in% 0:1), 0)
validate(logical(1) || (numeric(1) && . %in% 0:1), "1")
```

If you need to reference a literal dot (`.`) in a template, you can escape it by adding another dot so that `.` becomes `..`.  If you want to reference `...` you'll need to use `....`.  If you have a validation expression that doesn't reference the validation object (i.e. doesn't use `.`) you can mark it as a validation expression by wrapping it in `.()` (if you want to use a literal `.()` you can use `..()`).

If you need `&&` or `||` to be interpreted literally you can wrap the call in `identity` to tell `validate` not to recursively parse the expression:

```{r, eval=FALSE}
identity(length(a) == length(b) && . %in% 0:1)
```

which will be treated as a single custom expression instead of a template and a custom expression.

### Validation Tokens

#### Storing Validation Tokens in Variables

`validate` uses complex rules to parse and evaluate validation expressions, but they resolve to mostly intuitive outcomes.  One upside of the complexity is that you can actually store validation tokens or validation expressions in variables:

```{r}
TF <- quote(logical(1L) && !is.na(.))  # note `quote`
validate(TF, TRUE)
validate(TF, NA)
validate(TF, 1)
```

Since validate parses and evaluates tokens recursively, you can even combine pre-stored tokens:

```{r}
ZEROONE <- quote(numeric(1) && . %in% 0:1)
TF_ish <- quote(TF || ZEROONE)
validate(TF_ish, 1)
validate(TF_ish, "0")
```

You can also provide more helpful error messages with `mk_val_token`:

```{r}
TF <- mk_val_token(
  logical(1L) && !is.na(.), "be TRUE or FALSE" # `quote` not required
)
validate(TF, NA)
```
#### Validation Expression Parsing and Evaluation Rules

Hopefully you were able to get an intuitive sense of how validation expressions work.  If you are interested in the actual rules, read on, otherwise skip to the next section.

After parsing the validation expression, `validate` examines each token for the presence of `.` or `.()`.  If either are present then the `.` values are substituted for the corresponding variable name (`current` for `validate`, and whatever variable is matched to the validation expression for `validate_args`).  Then each token is evaluted in the parent frame.  If the token evaluated was a custom expression then it is checked for truth.  If not, the result of evaluating the token is used as a template for an `alike` check against the object being tested, except for the special case when the template evaluates to a quoted language object.  In that case `validate` recursively applies the same logic to the language object as described above.

#### Predefined Tokens

`validate` comes with many predifined validation tokens for common cases:

```{r}
validate(NUM.1.POS, 5)
validate(NUM.1.POS, -3)
validate(NUM.1.POS, runif(5))
validate(CHR, letters)
validate(CHR, factor(letters))
```

See `?validation_tokens` for more details about what predefined tokens are available.

## Performance Considerations

* Predifine templates
* benchmarks vs. `stopifnot`
* performance optimized for success; failure may be slower

## Argument Matching

## Acknowledgements

* Hadley for pointing me to smbache
* smbache for the obvious point that we might be interested in validating something other than function args, plus much easier to test
