---
title: "vetR - Trust, but Verify"
author: "Brodie Gaslam"
output:
    rmarkdown::html_vignette:
        toc: true
        css: styles.css

vignette: >
  %\VignetteIndexEntry{vetr}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(error=TRUE, comment=NA)
library(vetr)
```

```{r child='./rmdhunks/trust-but-verify.Rmd'}
```

```{r child='./rmdhunks/declarative-checks.Rmd'}
```

## Vetting Expressions

### Introduction

```{r child='./rmdhunks/vetting-expressions.Rmd'}
```

### Non Standard Evaluation

#### Vetting Expressions are Language Objects

`vet` captures the first argument unevaluated.  For example in:

```{r, eval=FALSE}
vet(. > 0, 1:3)
```

`. > 0` is captured, processed, and evaluated in a special manner.  This is a
common pattern in R (e.g. as in `with`, `subset`, etc.) called Non Standard
Evaluation (NSE).  One additional wrinkle with `vet` is that symbols in the
captured expression are recursively substituted:

```{r, eval=FALSE}
a <- quote(integer() && . > 0)
b <- quote(logical(1L) && !is.na(.))
c <- quote(a || b)

vet(c, 1:3)
```

The above is thus equivalent to:

```{r, eval=FALSE}
vet((integer() && . > 0) || (logical(1L) && !is.na(.)), 1:3)
```

The recursive substitution removes the typical limitation on "programming" with
NSE, although there are a few things to know:

* Symbols in vetting expressions that evaluate to language objects (calls or
  symbols) in the parent frame are substituted with the corresponding language
  object.
* The result of this substitution is implicitly wrapped in parentheses to avoid
  operator precedence problems.
* The function part of a call is never substituted (e.g. the `fun` in
  `fun(a, b)`); this extends to operators.
* `.` is never substituted, though you can work around that by escaping it with
  an additional `.` (i.e. `..`).
* You must take particular care when constructing vetting expressions for
  language objects.

To illustrate the last point, suppose we want to check that an object is a call
in the form `x + y`, then we could use:

```{r, eval=FALSE}
vet(quote(x + y), my.call)       # notice `quote`
```

Or:

```{r, eval=FALSE}
tpl.call <- quote(quote(x + y))  # notice `quote(quote(...))`
vet(tpl.call, my.call)
```

Additionally, you will need to ensure that `x` and `y` themselves do not
evaluate to language objects in the parent frame.

#### Parsing and Evaluation Rules

Once a vetting expression has been recursively substituted, it is parsed into
tokens.  Tokens are the parts of the vetting expression bounded by the `&&` and
`||` operators and optionally enclosed in parentheses.  For example, there are
three tokens in the following vetting expression:

```{r, eval=FALSE}
logical(1) || (numeric(1) && (. > 0 & . < 1))
```

They are `logical(1)`, `numeric(1)`, and `. > 0 & . < 1`.  The last
token is just one token not because of the parentheses around it but because it
is a call to `&` as opposed to `&&`.  Here we use the parentheses to remove
parsing ambiguity caused by `&` and `&&` having the same operator precedence.

After the tokens have been identified they are classified as standard tokens or
template tokens.  Standard tokens are those that contain the `.` symbol.  Every
other token is considered a template token.

Standard tokens are further processed by substituting any `.` with the value
of the object being vetted.  These tokens are then evaluated and if those
evaluations produce `TRUE` or logical vectors containing only `TRUE`, the tokens
pass, otherwise they fail.  With:

```{r}
vet(. > 0, 1:3)
```

`. > 0` becomes `1:3 > 0`, which evaluates to `c(TRUE, TRUE, TRUE)` and the
token passes.

Template tokens, i.e. tokens without a `.` symbol, are evaluated and the
resulting R object is sent along with the object to vet to `alike` for
structural comparison.  If `alike` returns `TRUE` then the token passes,
otherwise it fails.

Finally, the result of evaluating each token is plugged back into the original
expression.  So<sup>1</sup>:

```{r, eval=FALSE}
vet(logical(1) || (numeric(1) && (. > 0 & . < 1)), 42)
# becomes:
alike(logical(1L), 42) || (alike(numeric(1L), 42) && all(42 > 0 & 42 < 1))
# becomes:
FALSE || (TRUE && FALSE)
# becomes:
FALSE
```

And the vetting fails:

```{r}
vet(logical(1) || (numeric(1) && (. > 0 & . < 1)), 42)
```

### Special Cases

If you need to reference a literal dot (`.`) in a token, you can escape it by
adding another dot so that `.` becomes `..`.  If you want to reference `...`
you'll need to use `....`.  If you have a standard token that does not
reference the vetting object (i.e. does not use `.`) you can mark it as a
standard token by wrapping it in `.()` (if you want to use a literal `.()` you
can use `..()`).

If you need `&&` or `||` to be interpreted literally you can wrap the call in
`I` to tell `vet` to treat the entire call as a single token:

```{r, eval=FALSE}
I(length(a) == length(b) && . %in% 0:1)
```

`vet` will stop searching for tokens at the first call to a function other than
`(`, `&&`, and `||`.  The use of `I` here is just an example of this behavior
and convenient since `I` does not change the meaning of the vetting token.  An
implication of this is you should not nest template tokens inside functions as
`vet` will not identify them as template tokens and you may get unexpected
results.  For example:

```{r, eval=FALSE}
I(logical(1L) && my_special_fun(.))
```

will always fail because `logical(1L)` is part of a standard token and is
evaluated as `FALSE` rather than used a template token for a scalar logical.

## In Functions

The `vetr` function streamlines parameter checks in functions.  It behaves just
like `vet`, except that you need only specify the vetting expressions.  The
objects to vet are captured from the function environment:

```{r}
fun <- function(x, y, z) {
  vetr(
    matrix(numeric(), ncol=3),
    logical(1L),
    character(1L) && . %in% c("foo", "bar")
  )
  TRUE  # do work...
}
fun(matrix(1:12, 3), TRUE, "baz")
fun(matrix(1:12, 4), TRUE, "baz")
fun(matrix(1:12, 4), TRUE, "foo")
```

The arguments to `vetr` are matched to the arguments of the enclosing function
in the same way as with `match.call`.  For example, if we wished to vet just
the third argument:

```{r}
fun <- function(x, y, z) {
  vetr(z=character(1L) && . %in% c("foo", "bar"))
  TRUE  # do work...
}
fun(matrix(1:12, 3), TRUE, "baz")
fun(matrix(1:12, 4), TRUE, "bar")
```

Vetting expressions work the same way with `vetr` as they do with `vet`.

## Performance Considerations

### Benchmarks

`vetr` is written primarily in C to minimize the performance impact of adding
validation checks to your functions.  Performance should be faster than using
`stopifnot` except for the most trivial of checks.  The `vetr` function itself
carries some additional overhead from matching arguments, but it should still be
faster than `stopifnot` except in the simplest of cases.  Here we run our checks
on valid iris objects we used to illustrate [declarative
checks](#declarative-checks-with-templates):

```{r}
vetr_iris <- function(x) vetr(tpl.iris)

bench_mark(times=1e4,
  vet(tpl.iris, iris),
  vetr_iris(iris),
  stopifnot_iris(iris)   # defined in "Templates" section
)
```

Performance is optimized for the success case.  Failure cases should still
perform reasonably well, but will be slower than most success cases.

### Templates and Performance

Complex templates will be slower to evaluate than simple ones, particularly for
lists with lots of nested elements.  Note however that the cost of the vetting
expression is a function of the complexity of the template, not that of the
value being vetted.

We recommend that you predefine templates in your package and not in the
validation expression since some seemingly innocuous template creation
expressions carry substantial overhead:

```{r}
bench_mark(data.frame(a=numeric()))
```

In this case the `data.frame` call alone take over 100us.  In your package code
you could use:

```{r}
df.tpl <- data.frame(a=numeric())

my_fun <- function(x) {
  vetr(x=df.tpl)
  TRUE    # do work
}
```

This way the template is created once on package load and re-used each time your
function is called.

## Alternatives

There are many alternatives available to `vetr`.  We do a survey of the
following in our [parameter validation functions][4] review:

```{r child='./rmdhunks/related-packages.Rmd'}
```

----
<sup>1</sup>We take some liberties in this example for clarity.  For instance, `alike` returns a character vector on failure, not `FALSE`, so really what `vet` is doing is `isTRUE(alike(...))`.

[1]: ./vetr.html
[2]: ./alike.html
[3]: #non-standard-evaluation
[4]: http://htmlpreview.github.io/?https://github.com/brodieG/vetr/blob/master/extra/compare.html
